Ethikbumms*\\ {\footnotesize \textsuperscript{*}Mit bisschen Datenschutz bumms} \thanks{funding agency}

### 2022-05-27 draft.tex
public concerns in medical sector

\title{Ethikbumms*\\
{\footnotesize \textsuperscript{*}Mit bisschen Datenschutz bumms}
\thanks{Identify applicable funding agency here. If none, delete this.}
}

@@

\begin{abstract}
A few centuries ago, life love and a the universe were a mystery, but driven by deeply rooted human curiosity we are beginning to untangle the mess and understand the foundation that grew consciousness out of thin air. One piece of this puzzle is DesoxyriboNucleic Acid. The genetic material that after being refined by millenia of pure chance and survival holds the blueprint to every living creature on our planet. With growing understanding about its structure and the technological advancements that gave us tools to manipulate it, humanity has set its foot in the doorstep to a world in which DNA is no different than a LEGO brick and nothing is out of reach.
This paper aims to explore aspects of how this applied genetics will impact us in regard to ethics and data privacy. 
Genetic engineering is as much blessing as a curse and must be used with caution, while genome analysis unveils a tremendous amount of information not just about us, but about our relatives as well.
\end{abstract}
\section{Historical and Technical Introduction}

@@

\subsection{Medical Sector}
As preventing disease is usually preferable over treating it, a substantial amount of medicine focuses on early detection and prevention. Especially for the class of mendelian disorders, which are characterized by being linkable to a single gene defect, genetic testing has proven to make diagnosis easy and reliable \textbf{[citation needed]}.
However while a grand number of such diseases exist, they are usually rare so testing might only pay off if the defect is known to run in the family \textbf{[citation needed]}.
Much bigger potential lies in preventing common disease with either a genetic component or whose causes are spread over several genes and are therefore much harder to pinpoint. This uncertainty neccessarily introduces stochastics and the method known as genetic risk assessment \textbf{[citation needed]}.
Advances here have sprouted the field of pharmagenomics, which utilizes genetic information to practice precision medicine.
Just like ethnicity, genetic profile has been shown to affect drug response, thus using this knowledge to personalize prescriptions can increase effectiveness and lower cost and side effects \cite{costsaving}.
Predictive genetic testing does not come without concerns from both experts and the public.
Most prominent are patient autonomy and informed consent,genetic discrimation, commercial use, use of public resources towards genetic testing and research, patient responsibilities towards close relatives and psychological risks associated with carrying near certain information about the future\cite{publicopinion}.

\subsection{Insurance}
Insurance as a business walks a fine line between solidarity and preying on the vulnerable. Risk and how we manage it as a society is studied under the name of insurance ethics, but to date the literature is sparse.
While public insurance in practice is a form of wealth redistribution, private insurance is in essence paying for peace of mind\cite{insuranceethics}.
However this social safety net only functions as long as there is uncertainty, in fact it functions better the more uncertainty there is.
This becomes obvious in consideration of the idea that clarity about the future replaces the insurance sector.
In a fully deterministic world, omniscient private insurance companies would never lose money on a customer, as premiums could be precisely the cost said customer will produce over their contract, plus a small markup; in other words, not providing any insurance.
Keeping this in mind it seems logical that it's in the best interest of society at large to keep insurance companies mostly in the dark.
At least in the case of health insurance, this darkness is threatened by introducing genetesting aided risk analysis.
The inclusion of factors like genetic predisposition to disease in risk analysis is an incredibly helpful tool to doctors when choosing appropriate treatment and precautions, but becomes detrimental when accessed by insurance companies\cite{insurancegenes}.
Legal action to protect against this threat is already being taken, with the USA introducing the Genetic Information Nondiscrimination Act in 2008\cite{insurancegenes}.
It is vital to human development that we retain trust in and refine the methodology of genetic testing, while at the same time making sure the sensitive informations is kept discretely and genetic discrimination is kept under control.

>>>>

\title{Ethikbumms*\\
{\footnotesize \textsuperscript{*}Mit bisschen Datenschutz bumms}
\thanks{funding agency}
}

@@

\begin{abstract}
A few centuries ago, life love and a the universe were a mystery, but driven by deeply rooted human curiosity we are beginning to untangle the mess and understand the foundation that grew consciousness out of thin air. One piece of this puzzle is DesoxyriboNucleic Acid. The genetic material that after being refined by millenia of pure chance and survival holds the blueprint to every living creature on our planet. With growing understanding about its structure and the technological advancements that gave us tools to manipulate it, humanity has set its foot in the doorstep to a world in which DNA is no different than a LEGO brick and nothing is out of reach.
This paper aims to explore aspects of how this applied genetics will impact us in regard to ethics and data privacy. 
Since great responsibility comes with great power, Genetic engineering is as much blessing as curse and must be used with caution, while genome analysis unveils a tremendous amount of information not just about us, but about our relatives as well and enables more and more accurate predictions about the future.
\end{abstract}
\section{Historical and Technical Introduction}

@@

\subsection{Medical Sector}
As preventing disease is usually preferable over treating it, a substantial amount of medicine focuses on early detection and prevention. Especially for the class of mendelian disorders, which are characterized by being linkable to a single gene defect, genetic testing has proven to make diagnosis easy and reliable \textbf{[citation needed]}.
However while a grand number of such diseases exist, they are usually rare so testing might only pay off if the defect is known to run in the family \textbf{[citation needed]}.
Much bigger potential lies in preventing common disease with either a genetic component or whose causes are spread over several genes and are therefore much harder to pinpoint. This uncertainty neccessarily introduces stochastics and the method known as genetic risk assessment \textbf{[citation needed]}.
Advances here have sprouted the field of pharmagenomics, which utilizes genetic information to practice precision medicine.
Just like ethnicity, genetic profile has been shown to affect drug response, thus using this knowledge to personalize prescriptions can increase effectiveness and lower cost and side effects \cite{costsaving}.
Predictive genetic testing does not come without concerns from both experts and the public, the most prominent ones will be discussed in the following\cite{publicopinion}:

\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non negotiable, which neccessitates public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment.\cite{publicopinion}.
\subsubsection{Genetic Discrimation}
As discussed in the following chapter, subpar treatment grounded in genome analysis results is no better than discrimination on the basis of sex, race, age or religion.
\subsubsection{Commercial Use}
Profit as the primary incentive manages to corrupt even the most honourable endeavour.
Commercial gene testing and therapy raises the same questions about motive and ethical obligation to help that plagued the medical ethics for millenia.
Testing costs being inflated because of patents can hardly be in accordance with public interest, which sparks discussion about the next point\cite{publicopinion}.
\subsubsection{Use of Public Resources}
If private patents are to be avoided, public resources have to be used towards research in the applicable area.
Distributing those in a fair way however proves difficult. Developing tests is expensive so some conditions need to be prioritized, which in turn requires weighing of different factors. The initial cost of development is just as important as benefit it brings. This is comprised of the reliability of the result, as well as therapy options versus their respective cost. An unreliable result is equally useless from a public health perspective as knowledge about a genetic death sentence.
Additionally as many tests would only ever affect a small number of people, some argue that the money should either be used for more general genetic tests or towards programs that target public health in a broader sense like adressing diet\cite{publicopinion}.
\subsubsection{Patient Responsibilities}
While a general right to data privacy in medicine makes sense for traditional maladies, one might argue that since DNA test results have immediate consequences for relatives theres at least in part a responsibility to inform said relatives of any risks they themselves might face, comparable to infectious disease or STDs\cite{publicopinion}.
\subsubsection{Psychological Risks}
The psychological weight associated with carrying near certain information about the future is not to be underestimated in the context of genetic testing either.
Just as a cancer diagnosis, discovering to be at high risk of devoloping potentially lethal disease can be devastating, especially if treatment options are limited\cite{publicopinion}.

\subsection{Insurance}
Insurance as a business walks a fine line between solidarity and preying on the vulnerable. Risk and how we manage it as a society is studied under the name of insurance ethics, but to date the literature is sparse.
While public insurance in practice is a form of wealth redistribution, private insurance is in essence paying for peace of mind\cite{insuranceethics}.
However this social safety net only functions as long as there is uncertainty, in fact it functions better the more uncertainty there is.
This becomes obvious in consideration of the idea that clarity about the future replaces the insurance sector.
In a fully deterministic world, omniscient private insurance companies would never lose money on a customer, as premiums could be precisely the cost said customer will produce over their contract, plus a small markup; in other words, not providing any insurance.
Keeping this in mind it seems logical that it's in the best interest of society at large to keep insurance companies mostly in the dark.
At least in the case of health insurance, this darkness is threatened by introducing gene testing aided risk analysis.
The inclusion of factors like genetic predisposition to disease in risk analysis is an incredibly helpful tool to doctors when choosing appropriate treatment and taking precautions, but becomes detrimental when accessed by insurance companies\cite{insurancegenes}.
Legal action to protect against this threat is already being taken, with the USA introducing the Genetic Information Nondiscrimination Act in 2008\cite{insurancegenes}.
It is vital to human development that we retain trust in and refine the methodology of genetic testing, while at the same time making sure the sensitive informations is stored discretely and genetic discrimination is kept under control.
